Black-hole ringdown as a probe of higher-curvature gravity theories

### 2022-05-06 papers/paper_alt_theor_bounds/paper_alt_theor_bounds.tex
hs: cleaned commented out text / comments

% \usepackage{tikz}

@@

% HS: temporary title. Feel free to add suggestions
\title{Black-hole ringdown as a probe of higher-curvature gravity theories}

@@

These properties are not mere theoretical curiosities; they have important observational consequences.
%
First, the presence of the scalar charge implies that when in binaries, BHs can source
scalar-dipole radiation (see, e.g., Refs.~\cite{Yagi:2011xp,Julie:2019sab,Shiralilou:2020gah,Shiralilou:2021mfl,Julie:2022huo})
which affects the GW phase at $-1$PN order (relative to the dominant quadrupolar GR
contribution), with magnitude proportional to the difference between the
charges of binary components. This makes EdGB gravity testable with GW
observations of compact binaries where at least one component is a BH.
%
Indeed, Ref.~\cite{Lyu:2022gdr} placed the bound
%
% $\ell_{\rm EdGB} \leqslant 1.33$~km
$\ell_{\rm EdGB} \leqslant 7.1$~km
%
using the NSBH binaries GW200105 and
GW200115~\cite{LIGOScientific:2021qlt} while Refs.~\cite{Nair:2019iur,Perkins:2021mhb}, obtained
%
% $\ell_{\rm EdGB} \leqslant 1.7$~km,
$\ell_{\rm EdGB} \leqslant 9.1$~km,
%
by stacking the posteriors of $\ell_{\rm EdGB}$ from a selection of BBHs
from the GWTC-1 and GWTC-2 catalogs~\cite{LIGOScientific:2018mvr,LIGOScientific:2020ibl}\footnote{These bounds
are, strictly speaking, valid only when the scalar field $\varphi$ is small (i.e., $\varphi \ll 1$), and
they take into account only the leading-order scalar field interaction arising from the original dilatonic coupling
(i.e., $f(\varphi) \approx \varphi$ in Eq.~\eqref{eq:action_sgb}).
This results in what is often referred to as shift-symmetric scalar-Gauss-Bonnet theory.
In this theory, NSs do not have scalar monopole charge~\cite{Yagi:2015oca}, while BHs do~\cite{Yunes:2011we,Sotiriou:2013qea,Sotiriou:2014pfa}. Finally, we note that the constant $\alpha_{\rm EdGB}$ used in Refs.~\cite{Yagi:2011xp,Nair:2019iur,Perkins:2021mhb,Lyu:2022gdr} is related to $\ell_{\rm EdGB}$ as $\ell_{\rm EdGB} = 4 \pi^{1/4} |\alpha_{\rm EdGB}|^{1/2}$.}.
Secondly, the scalar field influences the response of BHs to linearized perturbations and thus
affects the BH's QNM spectra.
%
The coupling between scalar field and the Gauss-Bonnet invariant,
results in a coupling between scalar perturbations and gravitational
perturbations of polar parity which, for instance, breaks the equivalence
between the QNM spectra of polar and axial gravitational perturbations~\cite{Glampedakis:2017rar,Lenzi:2021njy}
(i.e., the isospectrality~\cite{Chandrasekhar:1985kt}) of Schwarzschild BHs in GR.

@@

This theory is described by the action~\cite{Jackiw:2003pm,Alexander:2009tp},
%
\begin{align} \label{eq:action_dcs}
    S_{\rm dCS} =
    \int \dV
    \left[
    \tfrac{1}{16\pi}\, R - \tfrac{1}{2}(\partial \vartheta)^2
    + \tfrac{1}{4} \ell^{2}_{\rm dCS} \, \vartheta \, {}^{*}RR
    \right],
\end{align}
%
where $\vartheta$ is a pseudoscalar field which couples to the Pontryagin
density
${}^{*}RR = {}^{*}R_{\mu\nu}{}^{\rho\sigma} R^{\nu\mu}{}_{\rho\sigma}$,
%
where ${}^{*}R^{\mu\nu\rho\sigma}$ is the dual of the Riemann tensor
defined as
%
${}^{*}R^{\mu\nu\rho\sigma} =
\epsilon^{\mu\nu}{}_{\gamma\delta}
R^{\gamma\delta\rho\sigma} / 2$,
%
and $\epsilon_{\mu\nu\gamma\delta}$ is the Levi-Civita tensor.
%
The variation of the Pontryagin density is a boundary
term that does not contribute to the field equations in
four-dimensions~\cite{Jackiw:2003pm}.
%
However, the Pontryagin density can modify the field equations when coupled to $\vartheta$;
the strength of the coupling set by $\ell_{\rm dCS}$, with dimensions of length.
%
% However, a coupling with $\vartheta$ can modify the field equations.
% %
% The coupling strength is set by the coupling constant $\ell_{\rm dCS}$, with
% dimensions of length.

@@

The GW signal from a quasicircular BBH can be described  in GR by a unique set of
parameters $\bm{\theta}$, that includes the masses and spins of the two BHs,
$(m_1,\, m_2,\, \mathbf{S}_1,\, \mathbf{S}_2)$, the sky location determined by the
luminosity distance $D_L$,
% \abcomm{I changed to capital $d$ because later we call the
% data ``d''.},
right ascension $\alpha$ and declination $\delta$,
and the orientation of the binary given by the inclination $\iota$ and polarization $\psi$
angles.
%
The set is completed by the choice of a reference time $t_0$ and phase
$\phi_0$. If we further assume that the spins of the individual BHs are
restricted to be aligned or anti-aligned (for short, aligned) to the unit vector
perpendicular to the orbital plane ($\mathbf{\hat{L}}$), we reduce the six
components of the spins to just two, $\chi_i \equiv \mathbf{S}_i\cdot \mathbf{\hat{L}} /m_i^2$ with $i = 1,2$,
and our entire parameter set from 15 to 11.
%
Let us also define some additional parameters and set some conventions that
will be useful in our analysis later, namely, the total mass
%
$M=m_1+m_2$,
%
the chirp mass
%
$\mathcal {M}=(m_{1}m_{2})^{3/5}/(m_{1}+m_{2})^{1/5}$,
%
the asymmetric mass ratio $q=m_1/m_2$, with the convention $m_1 \geqslant m_2$ (and thus $q \geqslant 1$),
and the symmetric mass ratio of the binary, $\nu = m_1m_2/(m_1+m_2)^2$. Note that for BHs $-1 \leqslant \chi_i \leqslant 1$.

For the polarizations of the GW signal (in the observer's frame) we have
%
\begin{equation}
h_+(\iota,\varphi_0;t ) - i h_\times(\iota,\varphi_0;t) = \sum_{l, m} {}_{-\!2}Y_{l m}(\iota,\varphi_0)\, h_{l m}(t),
\end{equation}
%
where ${}_{-\!2}Y_{l m}(\iota,\varphi_0)$ are the $-2$ spin-weighted spherical
harmonics.
%
% --------------------------------
% HS: maybe not necessary anymore:
% \footnote{Note that we use $l$ to denote the angular dependence index and $\ell$ to denote the coupling constant in Sec.~\ref{sec:review_parspec}.}
% --------------------------------
%
As our baseline model, that is, the GR model upon which non-GR modifications
are added, we use the computationally efficient (time-domain) multipolar
waveform model for quasicircular spin-aligned BBHs described in
Ref.~\cite{Mihaylov:2021bpf}, which contains the modes, $(l, |m|)=(2,2),(2,1)$, $(3,3)$, $(4,4)$,
and $(5,5)$. Such a model was built by applying the post-adiabatic approximation~\cite{Nagar:2018gnk}
to the multipolar spin-aligned EOB waveform model of Refs.~\cite{Bohe:2016gbl,Cotesta:2018fcv} (henceforth we refer to our baseline model as \SEOB{}\footnote{This waveform
model is available in \texttt{LALSuite}~\cite{lalsuite} as the \texttt{SEOBNRv4HM\_PA}
waveform approximant.}).

An accurate description of the merger is incorporated through calibration with
NR simulations, as described in Refs.~\cite{Bohe:2016gbl,Cotesta:2018fcv}, along with
information for the merger and ringdown phases, from BH perturbation theory.
The merger-ringdown waveform, $h_{l m}^\textrm{merger-RD}$, is then stitched to
inspiral-plunge waveform, $h_{l m}^\textrm{insp-plunge}$ at a certain time $t = t^{\textrm{match}}_{l m}$, as
%
\begin{equation}
h_{l m}(t) =
h_{l m}^\textrm{insp-plunge}\, \Theta(t_\textrm{match}^{l m} - t)
+ h_{l m}^\textrm{merger-RD}\,\Theta(t-t_\textrm{match}^{l m})\,,
\end{equation}
%
where $\Theta(t)$ is the Heaviside step function.
%
The merger-ringdown waveform is expressed as an exponentially damped
sinusoid~\cite{Bohe:2016gbl,Cotesta:2018fcv}
%
\begin{equation}
\label{RD}
h_{l m}^{\textrm{merger-RD}}(t) = \nu \ \tilde{A}_{l m}(t)\ e^{i \tilde{\phi}_{l m}(t)} \ e^{-i \sigma_{l m 0}(t-t^{\textrm{match}}_{l m})}\,,
\end{equation}
%
where
%
\begin{align}
\sigma_{l m 0} \equiv {\rm Re}(\sigma_{l m 0}) + i \, {\rm Im}(\sigma_{l m 0})
= \omega_{l m 0} - \frac{i}{\tau_{l m 0}}\,,
\end{align}
%
are the complex frequencies of the fundamental ($0$-th overtone) QNMs of the remnant BH.
%
%\ag{Going back to the convention established above to use the index $k$ to now denote the $(l, m,0)$
%mode,  the frequencies $\omega_k$ and damping times $\tau_k$ can be read off from the
%real and imaginary parts of $\sigma_k$ respectively.}
%
The functions $\tilde{A}_{l m}(t)$ and $\tilde{\phi}_{l m}(t)$ are defined
in Ref.~\cite{Bohe:2016gbl,Cotesta:2018fcv}.

@@

Using this parameterized waveform model, which we call $\pSEOB$,
we infer bounds on our non-GR parameter $\ell_{\rm th}$ for the specific cases of
modified gravity theories presented Sec.~\ref{sec:review_theories}.
%
We detail our results in Sec.~\ref{sec:results}.
%
% \hs{Alessandra, Abhirup, should we call this waveform differently to distinguish it from your previous papers?}
%
%In this paper, we will restrict ourselves to the $k=(2,2,0)$ QNM and corrections to the $j=0$ and $j=1$ terms, i.e., leading and next-to-leading order corrections to the spin, $\chi_f$. Assuming a prior which is uniform on the beyond-GR parameters, $(\ell, \{\delta \omega_k^{(j)}\},\{\delta \tau_k^{(j)}\})$, we employ a Bayesian framework to stochastically sample over the parameter space using the \texttt{LALInference} algorithm.~\cite{lallsuite}.
%
%The bounds on these beyond-GR parameters, for specific cases of modified gravity theories outlined in Sec.~\ref{sec:review_theories}, are detailed in Sec.~\ref{sec:results}.

@@

\begin{table*}[t]
% \begin{tabular}{c | c c c c c c c c c c}
% \hline
% \hline
% Theory      & $\omega^{(0)}_{220}$ & $\tau^{(0)}_{220}$ & $\omega^{(1)}_{220}$ & $\tau^{(1)}_{220}$ & $p$ & $\delta \omega^{(0)}_{220}$ & $\delta \tau^{(0)}_{220}$ & $\delta \omega^{(1)}_{220}$ & $\delta \tau^{(1)}_{220}$ & Ref.  \\
% \hline
% GR          & 0.3737               & 11.2407            &  0.1258              & 0.2522             & --  & --                          & --                        & --                          & --                        & \cite{Maselli:2019mjd} \\
% \hline
% EdGB        & --                   & --                 &  --                  & --                 & 4   & 0.0107                      & 0.0044                    & $-$0.2480                   & $-$1.101                  & \cite{Blazquez-Salcedo:2016enn,Pierini:2021jxd} \\
% dCS         & --                   & --                 &  --                  & --                 & 4   & 3.1964                      & 6.3619                    & 41.199                      & 794.66                    & \cite{Wagle:2021tam}  \\
% cubic EFT   & --                   & --                 &  --                  & --                 & 4   & $-$0.5813                   & 2.6469                    & $-$3.8620                   & 265.12                    & \cite{Cano:2021myl} \\
% quartic EFT & --                   & --                 &  --                  & --                 & 6   & $-$0.2114                   & $-$0.6070                 & $-$1.5263                   & 171.35                    & \cite{Cano:2021myl}  \\
% \hline
% \hline
% \end{tabular}
%
\begin{tabular}{c | c }
\hline
\hline
                        & GR~\cite{Maselli:2019mjd} \\
\hline
$\omega^{(0)}$    & 0.3737                    \\
$\tau^{(0)}$      & 11.2407                   \\
$\omega^{(1)}$    & 0.1258                    \\
$\tau^{(1)}$      & 0.2522                    \\
\hline
\hline
\end{tabular}
\quad
\begin{tabular}{c | c  c  c  c}
\hline
\hline
                              & EdGB ($p=4$)~\cite{Blazquez-Salcedo:2016enn,Pierini:2021jxd} & dCS ($p=4$)~\cite{Wagle:2021tam} & cubic EFT ($p=4$)~\cite{Cano:2021myl} & quartic EFT ($p=6$)~\cite{Cano:2021myl} \\
\hline
$\delta\omega^{(0)}$    & 0.0107                                               & 3.1964                   & $-$0.5813                     & $-$0.2114                       \\
$\delta\tau^{(0)}$      & 0.0044                                               & 6.3619                   & $-$0.2114                     & $-$0.6070                       \\
$\delta\omega^{(1)}$    & $-$0.2480                                            & 41.199                   & $-$3.8620                     & $-$1.5263                       \\
$\delta\tau^{(1)}$      & $-$1.1014                                            & 794.66                   & 265.12                        & 171.35                          \\
\hline
\hline
\end{tabular}
\caption{Summary of theory-specific QNM calculations.
%
We summarize each theory we consider together with: the exponent $p$ at
which their QNM-modification enters, the corresponding modifications to the
oscillation frequency $\delta \omega^{(n)}_{220}$ and decay time $\delta \tau^{(n)}_{220}$, and the
references from which we used the results from.
%
We also include for comparison the GR coefficients, up to $j=1$, obtained in Ref.~\cite{Maselli:2019mjd}.
%
The remaining GR coefficients for $1< j \leqslant 4$, for which their non-GR counterpart cannot be determined as of yet for
the theories under consideration, can be found in Table~I of Ref.~\cite{Maselli:2019mjd}.
%
}
\label{tab:ref_theories_qnms}
\end{table*}

@@

% \begin{table*}[th]
% \begin{tabular}{c | c c c c c c c c}
% \hline
% \hline
% Theory & $p$ & $\delta \omega^{(0)}_{220}$ & $\delta \tau^{(0)}_{220}$ & $\delta \omega^{(1)}_{220}$ & $\delta \tau^{(1)}_{220}$ & QNM & Constraint & This work \\
% \hline
% EdGB      & 4 & 0.0107 & 0.0044 & $-$0.2480 & $-$1.101 &  \cite{Blazquez-Salcedo:2016enn,Pierini:2021jxd} & $\ell_{\rm GB} \leqslant 1.18$~km~(GW)~\cite{Lyu:2022gdr} & -- \\
% dCS       & 4 & 3.1964 & 6.3619 & 41.199 & 794.66 &  \cite{Wagle:2021tam}   & $\ell_{\rm CS} \leqslant 8.5$~km~(EM+GW)~\cite{Silva:2020acr}  & $\ell_{\rm CS} \leqslant 34.5$~km \\
% cEFT  & 4 & -0.5813  & 2.6469 & -3.8620 & 265.12 & \cite{Cano:2021myl} & \dots  & \dots \\
% qEFT  & 6 & \dots  & \dots & \dots & \dots & \cite{Cano:2021myl} & $\ell_{\rm qEFT} = \dots$~km~\cite{Sennett:2019bpc}  & \dots \\
% \hline
% \hline
% \end{tabular}
% \caption{Summary of the quasinormal modes calculations.
% %
% We summarize each theory we have considered together with: the exponent $p$ at
% which their QNM-modification enters, the corresponding modifications to the
% oscillation frequency $\delta \omega^{(i)}_{220}$ and decay time $\delta \tau^{(i)}_{220}$, the
% references from which we used the results from and the current best constraint
% (if applicable).
% %
% In these cases, the dimensionless parameters $\lambda_{\rm e,o}$ are degenerate
% with the length scale $\ell$.
% %
% }
% \label{tab:ref_theories_qnms}
% \end{table*}

@@

\subsection{Bayesian formalism}
If we assume that a GW signal observed in detector data $d$ is accurately
described by our waveform model \pSEOB{}, we can infer the parameters of the
model, $\bm{\lambda}$, given the hypothesis $\mathcal{H}$, using Bayes' theorem,
%
\begin{equation}
P(\bm{\lambda} \vert d, \mathcal{H}) =
\frac{p(\bm{\lambda} \vert \mathcal{H}) \, \mathcal{L}(d \vert \bm{\lambda},\mathcal{H})}{E(d \vert \mathcal{H})}\,,
\label{eq:bayes}
\end{equation}
%
where $P(\bm{\lambda} \vert d, \mathcal{H})$ is the posterior probability distribution,
$p(\bm{\lambda} \vert \mathcal{H})$ the prior,
$\mathcal{L}(d \vert \bm{\lambda},\mathcal{H})$ the likelihood, and
$E(d \vert \mathcal{H})$ the evidence.
%
The set of parameters, $\bm{\lambda}$ is a union of the GR waveform
model parameters~$\bm{\theta}$ (see Sec.~\ref{sec:review_pSEOB})
% \abcomm{It seems to
% me that PRD does not use ``c.f.''. Does it? I am replacing it with ''see''.},
and $\ell_{\rm th}$, the only non-GR parameter in this problem which, we recall, sets
the characteristic length-scale in which deviations from
GR become relevant in each of the theories described in
Sec.~\ref{sec:review_theories}. %Hence,
%
%\begin{equation}
%\bm{\lambda} = \{\ell\} \cup \{\bm{\theta}\}.
%\label{eq:def_params}
%\end{equation}

% As discussed in Secs.~\ref{sec:review_theories} and \ref{sec:theory_specific_qnm}, there is a single coupling
% constants in the cases of EdGB and dCS theories, but two in the cubic and
% quartic EFTofGR.
% %
% We also hold the other beyond-GR parameters, $\{\delta \omega_k^{(j)}\},$ and $\{\delta \tau_k^{(j)}\}$
% fixed to theory-specific predictions. Hence, our set of beyond--GR parameters only includes the parameter $\{\ell \}$.
% %

@@

For $N$ independent GW observations $\{d_j\}$, $j=1,...,N$, each characterized
by a PDF $P_j(\ell_{\rm th} \vert d_j,\mathcal{H})$, the
joint posterior can be written as:
%
\begin{equation}
P(\ell_{\rm th} | \{d_j\},\mathcal{H}) = p(\ell_{\rm th}) \prod_{j=1}^{N} \frac{P_j(\ell_{\rm th} | d_j,\mathcal{H})}{p_j(\ell_{\rm th}|\mathcal{H})}\,.
\label{eq:cumulative_dist_ell}
\end{equation}
%
where $p_j(\ell_{\rm th} | \mathcal{H})$ are the priors used for each observation, \redsout{and} $p(\ell_{\rm th} | \mathcal{H})$ is an overall prior\hs{, and we assume that the value of $\ell_{\rm th}$ is shared among all events.}
%
Since we assume a uniform prior on $\ell_{\rm th}$, the joint posterior is
equal to the joint likelihood. Hereafter, we will drop the explicit usage of
$\mathcal{H}$.
%
%In some cases, we found it necessary to work with different prior ranges across
%different events, within the context of a given theory.
%
%In this situation, when combining the posteriors we use chose as our overall
%prior the widest one among the individual events.

@@

The prior distribution functions on the GR parameters are assumed to be
uniform over the component masses, $(m_1, m_2)$, isotropically distributed on a
sphere in the sky for the source location with $p(D_L) \propto D_L^2$, and isotropic
on the binary orientation, $p(\iota, \psi, \phi_0) \propto \sin\iota$. For the spins
$(\chi_1, \chi_2)$, we assume a prior uniform and isotropic in the spin magnitudes\footnote{This spin-prior choice can be specified in \texttt{LALInference} using the option \texttt{alignedspin-zprior.}}.
%
%The prior on $t_0$ is uniform and its range is informed by our detection pipelines.

Among our non-GR parameters $\{\ell_{\rm th}, \delta \omega^{(j)}\delta \tau^{(j)}\}$,
as already mentioned in the previous section, we hold $\{\delta \omega^{(j)},\delta \tau^{(j)}\}$
fixed to theory-specific predictions, and only allow $\ell_{\rm th}$ to vary freely.
%
We assume an uniform prior on $\ell_{\rm th}$ between appropriate ranges.
% \sout{, say $\ell_{\rm th} \in [0, \ell_{\rm max}]$}.
%
The lower limit is set by the fact that the modified gravity theories we consider
all have $p$ even and hence we can assume $\ell_{\rm th} > 0$ without loss of
generality.
%
%The upper limit is chosen as to make sure that our marginalized posterior
%distributions on $\ell$ do not rail against $\ell_{\rm max}$.

@@

% ----------------------------------------------------------
% \hs{Convention: us 90\% credibility for the final bounds.}
% ----------------------------------------------------------

% -----------------
% Explain threshold
% -----------------

>>>>

@@

\title{Black-hole ringdown as a probe of higher-curvature gravity theories}

@@

These properties are not mere theoretical curiosities; they have important observational consequences.
%
First, the presence of the scalar charge implies that when in binaries, BHs can source
scalar-dipole radiation (see, e.g., Refs.~\cite{Yagi:2011xp,Julie:2019sab,Shiralilou:2020gah,Shiralilou:2021mfl,Julie:2022huo})
which affects the GW phase at $-1$PN order (relative to the dominant quadrupolar GR
contribution), with magnitude proportional to the difference between the
charges of binary components. This makes EdGB gravity testable with GW
observations of compact binaries where at least one component is a BH.
%
Indeed, Ref.~\cite{Lyu:2022gdr} placed the bound
%
$\ell_{\rm EdGB} \leqslant 7.1$~km
%
using the NSBH binaries GW200105 and
GW200115~\cite{LIGOScientific:2021qlt} while Refs.~\cite{Nair:2019iur,Perkins:2021mhb}, obtained
%
$\ell_{\rm EdGB} \leqslant 9.1$~km,
%
by stacking the posteriors of $\ell_{\rm EdGB}$ from a selection of BBHs
from the GWTC-1 and GWTC-2 catalogs~\cite{LIGOScientific:2018mvr,LIGOScientific:2020ibl}\footnote{These bounds
are, strictly speaking, valid only when the scalar field $\varphi$ is small (i.e., $\varphi \ll 1$), and
they take into account only the leading-order scalar field interaction arising from the original dilatonic coupling
(i.e., $f(\varphi) \approx \varphi$ in Eq.~\eqref{eq:action_sgb}).
This results in what is often referred to as shift-symmetric scalar-Gauss-Bonnet theory.
In this theory, NSs do not have scalar monopole charge~\cite{Yagi:2015oca}, while BHs do~\cite{Yunes:2011we,Sotiriou:2013qea,Sotiriou:2014pfa}. Finally, we note that the constant $\alpha_{\rm EdGB}$ used in Refs.~\cite{Yagi:2011xp,Nair:2019iur,Perkins:2021mhb,Lyu:2022gdr} is related to $\ell_{\rm EdGB}$ as $\ell_{\rm EdGB} = 4 \pi^{1/4} |\alpha_{\rm EdGB}|^{1/2}$.}.
Secondly, the scalar field influences the response of BHs to linearized perturbations and thus
affects the BH's QNM spectra.
%
The coupling between scalar field and the Gauss-Bonnet invariant,
results in a coupling between scalar perturbations and gravitational
perturbations of polar parity which, for instance, breaks the equivalence
between the QNM spectra of polar and axial gravitational perturbations~\cite{Glampedakis:2017rar,Lenzi:2021njy}
(i.e., the isospectrality~\cite{Chandrasekhar:1985kt}) of Schwarzschild BHs in GR.

@@

This theory is described by the action~\cite{Jackiw:2003pm,Alexander:2009tp},
%
\begin{align} \label{eq:action_dcs}
    S_{\rm dCS} =
    \int \dV
    \left[
    \tfrac{1}{16\pi}\, R - \tfrac{1}{2}(\partial \vartheta)^2
    + \tfrac{1}{4} \ell^{2}_{\rm dCS} \, \vartheta \, {}^{*}RR
    \right],
\end{align}
%
where $\vartheta$ is a pseudoscalar field which couples to the Pontryagin
density
${}^{*}RR = {}^{*}R_{\mu\nu}{}^{\rho\sigma} R^{\nu\mu}{}_{\rho\sigma}$,
%
where ${}^{*}R^{\mu\nu\rho\sigma}$ is the dual of the Riemann tensor
defined as
%
${}^{*}R^{\mu\nu\rho\sigma} =
\epsilon^{\mu\nu}{}_{\gamma\delta}
R^{\gamma\delta\rho\sigma} / 2$,
%
and $\epsilon_{\mu\nu\gamma\delta}$ is the Levi-Civita tensor.
%
The variation of the Pontryagin density is a boundary
term that does not contribute to the field equations in
four-dimensions~\cite{Jackiw:2003pm}.
%
However, the Pontryagin density can modify the field equations when coupled to $\vartheta$;
the strength of the coupling set by $\ell_{\rm dCS}$, with dimensions of length.

@@

The GW signal from a quasicircular BBH can be described  in GR by a unique set of
parameters $\bm{\theta}$, that includes the masses and spins of the two BHs,
$(m_1,\, m_2,\, \mathbf{S}_1,\, \mathbf{S}_2)$, the sky location determined by the
luminosity distance $D_L$,
right ascension $\alpha$ and declination $\delta$,
and the orientation of the binary given by the inclination $\iota$ and polarization $\psi$
angles.
%
The set is completed by the choice of a reference time $t_0$ and phase
$\phi_0$. If we further assume that the spins of the individual BHs are
restricted to be aligned or anti-aligned (for short, aligned) to the unit vector
perpendicular to the orbital plane ($\mathbf{\hat{L}}$), we reduce the six
components of the spins to just two, $\chi_i \equiv \mathbf{S}_i\cdot \mathbf{\hat{L}} /m_i^2$ with $i = 1,2$,
and our entire parameter set from 15 to 11.
%
Let us also define some additional parameters and set some conventions that
will be useful in our analysis later, namely, the total mass
%
$M=m_1+m_2$,
%
the chirp mass
%
$\mathcal {M}=(m_{1}m_{2})^{3/5}/(m_{1}+m_{2})^{1/5}$,
%
the asymmetric mass ratio $q=m_1/m_2$, with the convention $m_1 \geqslant m_2$ (and thus $q \geqslant 1$),
and the symmetric mass ratio of the binary, $\nu = m_1m_2/(m_1+m_2)^2$. Note that for BHs $-1 \leqslant \chi_i \leqslant 1$.

For the polarizations of the GW signal (in the observer's frame) we have
%
\begin{equation}
h_+(\iota,\varphi_0;t ) - i h_\times(\iota,\varphi_0;t) = \sum_{l, m} {}_{-\!2}Y_{l m}(\iota,\varphi_0)\, h_{l m}(t),
\end{equation}
%
where ${}_{-\!2}Y_{l m}(\iota,\varphi_0)$ are the $-2$ spin-weighted spherical
harmonics.
%
As our baseline model, that is, the GR model upon which non-GR modifications
are added, we use the computationally efficient (time-domain) multipolar
waveform model for quasicircular spin-aligned BBHs described in
Ref.~\cite{Mihaylov:2021bpf}, which contains the modes, $(l, |m|)=(2,2),(2,1)$, $(3,3)$, $(4,4)$,
and $(5,5)$. Such a model was built by applying the post-adiabatic approximation~\cite{Nagar:2018gnk}
to the multipolar spin-aligned EOB waveform model of Refs.~\cite{Bohe:2016gbl,Cotesta:2018fcv} (henceforth we refer to our baseline model as \SEOB{}\footnote{This waveform
model is available in \texttt{LALSuite}~\cite{lalsuite} as the \texttt{SEOBNRv4HM\_PA}
waveform approximant.}).

An accurate description of the merger is incorporated through calibration with
NR simulations, as described in Refs.~\cite{Bohe:2016gbl,Cotesta:2018fcv}, along with
information for the merger and ringdown phases, from BH perturbation theory.
The merger-ringdown waveform, $h_{l m}^\textrm{merger-RD}$, is then stitched to
inspiral-plunge waveform, $h_{l m}^\textrm{insp-plunge}$ at a certain time $t = t^{\textrm{match}}_{l m}$, as
%
\begin{equation}
h_{l m}(t) =
h_{l m}^\textrm{insp-plunge}\, \Theta(t_\textrm{match}^{l m} - t)
+ h_{l m}^\textrm{merger-RD}\,\Theta(t-t_\textrm{match}^{l m})\,,
\end{equation}
%
where $\Theta(t)$ is the Heaviside step function.
%
The merger-ringdown waveform is expressed as an exponentially damped
sinusoid~\cite{Bohe:2016gbl,Cotesta:2018fcv}
%
\begin{equation}
\label{RD}
h_{l m}^{\textrm{merger-RD}}(t) = \nu \ \tilde{A}_{l m}(t)\ e^{i \tilde{\phi}_{l m}(t)} \ e^{-i \sigma_{l m 0}(t-t^{\textrm{match}}_{l m})}\,,
\end{equation}
%
where
%
\begin{align}
\sigma_{l m 0} \equiv {\rm Re}(\sigma_{l m 0}) + i \, {\rm Im}(\sigma_{l m 0})
= \omega_{l m 0} - \frac{i}{\tau_{l m 0}}\,,
\end{align}
%
are the complex frequencies of the fundamental ($0$-th overtone) QNMs of the remnant BH.
%
The functions $\tilde{A}_{l m}(t)$ and $\tilde{\phi}_{l m}(t)$ are defined
in Ref.~\cite{Bohe:2016gbl,Cotesta:2018fcv}.

@@

Using this parameterized waveform model, which we call $\pSEOB$,
we infer bounds on our non-GR parameter $\ell_{\rm th}$ for the specific cases of
modified gravity theories presented Sec.~\ref{sec:review_theories}.
%
We detail our results in Sec.~\ref{sec:results}.

@@

\begin{table*}[t]
\begin{tabular}{c | c }
\hline
\hline
                        & GR~\cite{Maselli:2019mjd} \\
\hline
$\omega^{(0)}$    & 0.3737                    \\
$\tau^{(0)}$      & 11.2407                   \\
$\omega^{(1)}$    & 0.1258                    \\
$\tau^{(1)}$      & 0.2522                    \\
\hline
\hline
\end{tabular}
\quad
\begin{tabular}{c | c  c  c  c}
\hline
\hline
                              & EdGB ($p=4$)~\cite{Blazquez-Salcedo:2016enn,Pierini:2021jxd} & dCS ($p=4$)~\cite{Wagle:2021tam} & cubic EFT ($p=4$)~\cite{Cano:2021myl} & quartic EFT ($p=6$)~\cite{Cano:2021myl} \\
\hline
$\delta\omega^{(0)}$    & 0.0107                                               & 3.1964                   & $-$0.5813                     & $-$0.2114                       \\
$\delta\tau^{(0)}$      & 0.0044                                               & 6.3619                   & $-$0.2114                     & $-$0.6070                       \\
$\delta\omega^{(1)}$    & $-$0.2480                                            & 41.199                   & $-$3.8620                     & $-$1.5263                       \\
$\delta\tau^{(1)}$      & $-$1.1014                                            & 794.66                   & 265.12                        & 171.35                          \\
\hline
\hline
\end{tabular}
\caption{Summary of theory-specific QNM calculations.
%
We summarize each theory we consider together with: the exponent $p$ at
which their QNM-modification enters, the corresponding modifications to the
oscillation frequency $\delta \omega^{(n)}_{220}$ and decay time $\delta \tau^{(n)}_{220}$, and the
references from which we used the results from.
%
We also include for comparison the GR coefficients, up to $j=1$, obtained in Ref.~\cite{Maselli:2019mjd}.
%
The remaining GR coefficients for $1< j \leqslant 4$, for which their non-GR counterpart cannot be determined as of yet for
the theories under consideration, can be found in Table~I of Ref.~\cite{Maselli:2019mjd}.
%
}
\label{tab:ref_theories_qnms}
\end{table*}

@@

@@

\subsection{Bayesian formalism}
If we assume that a GW signal observed in detector data $d$ is accurately
described by our waveform model \pSEOB{}, we can infer the parameters of the
model, $\bm{\lambda}$, given the hypothesis $\mathcal{H}$, using Bayes' theorem,
%
\begin{equation}
P(\bm{\lambda} \vert d, \mathcal{H}) =
\frac{p(\bm{\lambda} \vert \mathcal{H}) \, \mathcal{L}(d \vert \bm{\lambda},\mathcal{H})}{E(d \vert \mathcal{H})}\,,
\label{eq:bayes}
\end{equation}
%
where $P(\bm{\lambda} \vert d, \mathcal{H})$ is the posterior probability distribution,
$p(\bm{\lambda} \vert \mathcal{H})$ the prior,
$\mathcal{L}(d \vert \bm{\lambda},\mathcal{H})$ the likelihood, and
$E(d \vert \mathcal{H})$ the evidence.
%
The set of parameters, $\bm{\lambda}$ is a union of the GR waveform
model parameters~$\bm{\theta}$ (see Sec.~\ref{sec:review_pSEOB})
% \abcomm{It seems to
% me that PRD does not use ``c.f.''. Does it? I am replacing it with ''see''.},
and $\ell_{\rm th}$, the only non-GR parameter in this problem which, we recall, sets
the characteristic length-scale in which deviations from
GR become relevant in each of the theories described in
Sec.~\ref{sec:review_theories}.

@@

For $N$ independent GW observations $\{d_j\}$, $j=1,...,N$, each characterized
by a PDF $P_j(\ell_{\rm th} \vert d_j,\mathcal{H})$, the
joint posterior can be written as:
%
\begin{equation}
P(\ell_{\rm th} | \{d_j\},\mathcal{H}) = p(\ell_{\rm th}) \prod_{j=1}^{N} \frac{P_j(\ell_{\rm th} | d_j,\mathcal{H})}{p_j(\ell_{\rm th}|\mathcal{H})}\,.
\label{eq:cumulative_dist_ell}
\end{equation}
%
where $p_j(\ell_{\rm th} | \mathcal{H})$ are the priors used for each observation, \redsout{and} $p(\ell_{\rm th} | \mathcal{H})$ is an overall prior\hs{, and we assume that the value of $\ell_{\rm th}$ is shared among all events.}
%
Since we assume a uniform prior on $\ell_{\rm th}$, the joint posterior is
equal to the joint likelihood. Hereafter, we will drop the explicit usage of
$\mathcal{H}$.

@@

The prior distribution functions on the GR parameters are assumed to be
uniform over the component masses, $(m_1, m_2)$, isotropically distributed on a
sphere in the sky for the source location with $p(D_L) \propto D_L^2$, and isotropic
on the binary orientation, $p(\iota, \psi, \phi_0) \propto \sin\iota$. For the spins
$(\chi_1, \chi_2)$, we assume a prior uniform and isotropic in the spin magnitudes\footnote{This spin-prior choice can be specified in \texttt{LALInference} using the option \texttt{alignedspin-zprior.}}.

Among our non-GR parameters $\{\ell_{\rm th}, \delta \omega^{(j)}\delta \tau^{(j)}\}$,
as already mentioned in the previous section, we hold $\{\delta \omega^{(j)},\delta \tau^{(j)}\}$
fixed to theory-specific predictions, and only allow $\ell_{\rm th}$ to vary freely.
%
We assume an uniform prior on $\ell_{\rm th}$ between appropriate ranges.
%
The lower limit is set by the fact that the modified gravity theories we consider
all have $p$ even and hence we can assume $\ell_{\rm th} > 0$ without loss of
generality.

@@
